---
title: "zealgt vs zealbc1: why the variants differ"
subtitle: "Donor Zx.0540_P3, chromosome 10 (2026-10-05)"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-width: 8
    fig-height: 4.5
execute:
  echo: false
  warning: false
  message: false
---

zealgt found the high-confidence variants of donor Zx.0540_P3 on chromosome 10, from the same 5 BC1 samples and the
same lines that zealbc1 used. We expected the two sets to overlap almost fully (99 %). They overlapped 60 %. Overlap
here means shared variants divided by all variants found by either.

::: {.callout-note}
## In short

- **Our variant calling matches zealbc1's.** Given the same reads, both call exactly the same variants, and variants
  found by both get similar scores.
- **The difference is in the reads: PCR duplicates.** zealbc1 counted copies of the same DNA fragment as independent
  evidence; zealgt counts each fragment once. With duplicates counted, zealgt finds 30 % more high-confidence variants
  and the overlap rises to 72 %.
- **zealgt keeps removing duplicates.** The remaining difference comes from how the reads were cleaned before
  alignment. Not pursued: it would compare two cleaning recipes, not accuracy.
:::

**Confidence levels.** Each variant gets a score: how much better "real variant" explains its reads than "sequencing
error", as odds on a log scale (zealbc1 and zealgt use the same rules).

| Label here | Pipeline tier | Rule |
|---|---|---|
| high | A | odds ≥ 1,000 : 1 (score ≥ 6.9), variant reads in ≥ 2 BC1 samples, no warning |
| medium | B | odds ≥ 100 : 1 (score ≥ 4.6), no depth warning |
| low | C | odds ≥ 9 : 1 (score ≥ 2.2) |
| B73 allele | ref | odds ≥ 55 : 1 for error (score ≤ −4), ≥ 12 reads |
| rejected | none | anything else |

Warnings: *too deep* (over twice the usual read depth, a likely repeat), *over half variant reads* (more than a BC1
sample can carry), *contradictory* (one BC1 sample with ≥ 3 variant reads while another has ≥ 10 reads and none).

```{r setup}
library(data.table)
library(ggplot2)
library(patchwork)

d <- "genotype_pilot_chr10"
sites <- fread(file.path(d, "sites.tsv.gz"))
sites[, key := paste(pos, ref, alt)]
theme_set(theme_minimal(base_size = 11) + theme(panel.grid.minor = element_blank()))
pal <- c(shared = "#4C78A8", `zealgt only` = "#F58518", `zealbc1 only` = "#54A24B")
run_label <- c(
  zealbc1 = "zealbc1", zealgt_mbq20 = "zealgt",
  zealgt_mbq10 = "zealgt, lower base-quality cutoff", zealgt_nodup = "zealgt, duplicates counted"
)
tier_label <- c(A = "high", B = "medium", C = "low", ref = "B73 allele", none = "rejected")

high <- function(run_name) sites[run == run_name & tier == "A" & in_both_beds == 1, pos]
overlap <- function(x, y) {
  a <- high(x)
  b <- high(y)
  data.table(
    shared = length(intersect(a, b)), x_only = length(setdiff(a, b)), y_only = length(setdiff(b, a)),
    jaccard = length(intersect(a, b)) / length(union(a, b))
  )
}
```

## 1. How much the variant sets overlap

High-confidence variants under each test. The top three bars compare zealgt with zealbc1; the bottom two compare zealgt
with itself.

```{r fig-overlap, fig.height = 4}
pairs <- list(
  c("zealgt_mbq20", "zealbc1", "zealgt vs zealbc1"),
  c("zealgt_mbq10", "zealbc1", "lower base-quality cutoff vs zealbc1"),
  c("zealgt_nodup", "zealbc1", "duplicates counted vs zealbc1"),
  c("zealgt_mbq10", "zealgt_mbq20", "zealgt: lower vs usual base-quality cutoff"),
  c("zealgt_nodup", "zealgt_mbq20", "zealgt: duplicates counted vs removed")
)
ov <- rbindlist(lapply(pairs, function(p) overlap(p[1], p[2])[, comparison := p[3]]))
ov[, comparison := factor(comparison, rev(comparison))]
long <- melt(ov, id.vars = c("comparison", "jaccard"), measure.vars = c("shared", "x_only", "y_only"))
long[, part := fcase(variable == "shared", "shared", variable == "x_only", "zealgt only", default = "zealbc1 only")]
ggplot(long, aes(value / 1000, comparison, fill = part)) +
  geom_col(width = 0.7) +
  geom_text(
    data = ov, aes(x = (shared + x_only + y_only) / 1000, y = comparison, label = sprintf("%.0f %%", 100 * jaccard)),
    inherit.aes = FALSE, hjust = -0.15, size = 3.5
  ) +
  scale_fill_manual(values = pal, labels = c(
    shared = "in both", `zealgt only` = "first only", `zealbc1 only` = "second only"
  )) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "high-confidence variants (thousands)", y = NULL, fill = NULL) +
  theme(legend.position = "bottom")
```

Lowering the base-quality cutoff to zealbc1's barely changes anything (99 % overlap with our usual run). Counting
duplicates changes a lot, and it is the only change that brings us closer to zealbc1.

## 2. Where the variants are lost

The variant caller (CRISP) runs first; then variants without support in the pooled lines are dropped; then each
variant is scored.

```{r fig-stages, fig.height = 4}
stg <- fread(file.path(d, "stages.tsv"))
stage_levels <- c("CRISP, in BED", "after the witness veto", "site table", "tier A")
stage_labels <- c("called", "supported\nby the lines", "scored\n(all levels)", "high")
stg[, stage := factor(stage, levels = stage_levels, labels = stage_labels)]
p1 <- ggplot(stg, aes(stage, records / 1000, fill = pipeline)) +
  geom_col(position = position_dodge(0.75), width = 0.7) +
  scale_fill_manual(values = c(zealbc1 = "#54A24B", zealgt = "#F58518")) +
  labs(x = NULL, y = "variants (thousands)", fill = NULL) +
  theme(legend.position = "top")
tr <- fread(file.path(d, "absent_trace.tsv"))
tr[, direction := fifelse(
  grepl("^zealbc1", direction), "zealbc1's variants\nmissing in zealgt", "zealgt's variants\nmissing in zealbc1"
)]
where_levels <- c("not called by CRISP", "removed by the witness veto", "kept by the veto, not scored")
where_labels <- c("never called", "no support in the lines", "not a simple SNP in zealbc1")
tr[, where := factor(where, levels = where_levels, labels = where_labels)]
p2 <- ggplot(tr, aes(sites / 1000, direction, fill = where)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c("#9D755D", "#BAB0AC", "#B279A2")) +
  labs(x = "high-confidence variants (thousands)", y = NULL, fill = NULL) +
  theme(legend.position = "top", legend.direction = "vertical")
p1 + p2 + plot_layout(widths = c(1, 1.2))
```

The loss is at the very first step: our caller finds a third fewer variants. The later filters are not the cause.
Only simple SNPs (one base, two alleles) are scored; zealgt drops the others together with the line filter,
zealbc1 just before scoring.

## 3. Variants found by both get the same verdict

```{r fig-tiers, fig.height = 4.2}
both_tables <- function(x) {
  m <- merge(sites[run == x, .(key, tier_x = tier, llr_x = LLR)],
    sites[run == "zealbc1", .(key, tier_y = tier, llr_y = LLR)],
    by = "key"
  )
  m[, run := run_label[x]]
}
bt <- rbind(both_tables("zealgt_mbq20"), both_tables("zealgt_nodup"))
xt <- bt[, .N, by = .(run, tier_x, tier_y)]
xt[, `:=`(
  tier_x = factor(tier_label[tier_x], rev(tier_label)),
  tier_y = factor(tier_label[tier_y], tier_label)
)]
ggplot(xt, aes(tier_y, tier_x, fill = log10(N))) +
  geom_tile(colour = "white") +
  geom_text(aes(label = format(N, big.mark = ",")), size = 3) +
  facet_wrap(~run) +
  scale_fill_gradient(low = "#EEF3F9", high = "#4C78A8", guide = "none") +
  labs(x = "confidence in zealbc1", y = "confidence in zealgt") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r fig-llr, fig.height = 4.2}
cors <- bt[, .(r = cor(llr_x, llr_y)), by = run]
ggplot(bt, aes(llr_y, llr_x)) +
  geom_bin2d(bins = 70) +
  geom_abline(linetype = 2, colour = "grey40") +
  geom_text(data = cors, aes(x = 5, y = 230, label = sprintf("r = %.2f", r)), hjust = 0, inherit.aes = FALSE) +
  facet_wrap(~run) +
  scale_fill_viridis_c(trans = "log10") +
  coord_cartesian(xlim = c(-10, 250), ylim = c(-10, 250)) +
  labs(x = "zealbc1 score", y = "zealgt score", fill = "variants")
```

Most variants sit on the diagonal: the two pipelines disagree about which variants exist, not about how good they are.
The scores also show the duplicate inflation. With duplicates removed, zealgt's scores run about 7 % below zealbc1's
(slope 0.93); with duplicates counted, they run about 9 % above (slope 1.09), the excess coming from zealgt's slightly
deeper reads. Counting copies as evidence raises every score.

## 4. Swapping one input at a time

The caller on the first 10 Mb of chromosome 10, with zealbc1's settings, changing one input per run. Percentages are the
overlap with run 1.

```{r fig-grid, fig.height = 3.6}
grid <- fread(file.path(d, "grid.tsv"))
grid[, label := c(
  "1: our caller, zealbc1's reads",
  "2: zealbc1's caller, zealbc1's reads",
  "3: our caller, our lines",
  "4: our caller, our BC1 reads",
  "5: our caller, our BC1 reads, duplicates counted"
)[match(job, c(1, 2, 4, 3, 5))]]
grid[, label := factor(label, rev(label))]
ggplot(grid, aes(positions, label)) +
  geom_col(fill = "#4C78A8", width = 0.6) +
  geom_text(aes(label = sprintf("%s  (%.0f %%)", format(positions, big.mark = ","), 100 * jaccard_with_job1)),
    hjust = -0.05, size = 3.5
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.25))) +
  labs(x = "variants called", y = NULL)
```

- Our caller and zealbc1's give identical results (runs 1, 2).
- Our pooled lines change the result a little (run 3).
- Our BC1 reads cut the variants by a third (run 4); counting duplicates brings the number back (run 5).

## 5. The duplicates

Reads covering 693 variants that only zealbc1 found, and 358 that both found, first 10 Mb.

```{r fig-reads, fig.height = 4.2}
rf <- fread(file.path(d, "read_features.tsv"))[feature %in% c("near_5p", "near_3p", "soft_clipped", "coord_dup")]
feature_labels <- c(
  soft_clipped = "variant in a clipped part", near_5p = "variant near the read start",
  near_3p = "variant near the read end", coord_dup = "duplicate read"
)
rf[, feature := factor(feature_labels[feature], feature_labels)]
rf[, sites := fifelse(sites == "shared", "found by both", "found by zealbc1 only")]
p1 <- ggplot(rf, aes(fraction, feature, colour = crams, shape = sites)) +
  geom_point(size = 2.6, position = position_dodge(0.5)) +
  scale_colour_manual(values = c(zealbc1 = "#54A24B", zealgt = "#F58518")) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "share of the reads carrying the variant", y = NULL, colour = "reads of", shape = NULL)
rs <- fread(file.path(d, "read_sites.tsv"))[crams == "zealgt"]
rs <- melt(rs, id.vars = c("sites", "pos"), measure.vars = c("alt_reads", "alt_reads_passing"))
rs[, variable := fifelse(variable == "alt_reads", "all reads", "duplicates removed")]
rs[, sites := fifelse(sites == "shared", "found by both", "found by zealbc1 only")]
p2 <- ggplot(rs, aes(variable, value, fill = sites)) +
  geom_boxplot(outlier.size = 0.3, width = 0.6, position = position_dodge(0.7)) +
  scale_fill_manual(values = c(`found by both` = "#4C78A8", `found by zealbc1 only` = "#54A24B")) +
  coord_cartesian(ylim = c(0, 40)) +
  labs(x = NULL, y = "reads carrying the variant", fill = NULL)
p1 / p2 + plot_layout(heights = c(1, 1))
```

Read ends and clipping look the same in both pipelines, so they are not the cause. Duplicates are: variants that only
zealbc1 found lean more on duplicate reads (30 % vs 21 %), and once duplicates are removed they have less support.

## 6. Counting duplicates on the whole chromosome

```{r fig-nodup, fig.height = 3.6}
tc <- sites[run %in% c("zealgt_mbq20", "zealgt_nodup", "zealbc1") & in_both_beds == 1, .N, by = .(run, tier)]
tc[, `:=`(
  run = factor(run_label[run], run_label[c("zealgt_mbq20", "zealgt_nodup", "zealbc1")]),
  tier = factor(tier_label[tier], tier_label)
)]
ggplot(tc, aes(tier, N / 1000, fill = run)) +
  geom_col(position = position_dodge(0.8), width = 0.75) +
  scale_fill_manual(values = c("#F58518", "#E45756", "#54A24B")) +
  labs(x = "confidence", y = "variants (thousands)", fill = NULL) +
  theme(legend.position = "top")
```

Counting duplicates raises our high-confidence variants from 32,749 to 42,934, close to zealbc1's 40,941. So about a
quarter of zealbc1's high-confidence variants rest on duplicate reads.

```{r fig-position, fig.height = 3.2}
ours <- high("zealgt_mbq20")
pos_a <- sites[run == "zealbc1" & tier == "A" & in_both_beds == 1]
pos_a[, group := fifelse(pos %in% ours, "found by both", "found by zealbc1 only")]
ggplot(pos_a, aes(pos / 1e6, fill = group)) +
  geom_histogram(binwidth = 2, position = "identity", alpha = 0.7) +
  scale_fill_manual(values = c(`found by both` = "#4C78A8", `found by zealbc1 only` = "#54A24B")) +
  labs(x = "chromosome 10 (Mb)", y = "high-confidence variants", fill = NULL)
```

The missing variants are spread along the whole chromosome, not clustered in one region.

## Not pursued

With duplicates counted, 28 % of the variants still differ. That comes from how each pipeline cleaned the reads before
alignment (barcode matching, adapter trimming, mapping-quality filter). Testing each step means realigning all 5 BC1
samples per step, and would only compare two cleaning recipes. Accuracy will be measured on simulated data instead.

## Conclusion

zealgt goes forward as it is (user, 2026-10-05):

- **Duplicates are marked.** The score treats each read as independent evidence; a PCR copy repeats one fragment and
  adds nothing new. About a quarter of zealbc1's high-confidence variants rest on such copies (sections 5 and 6).
- **Reads are processed as Twist documents for the kits.** It gives slightly more reads (0-10 % more depth at the
  variants both pipelines find) and no sign of artifacts (section 5: read ends and clipped parts look the same in both
  pipelines). It is also the method that follows the manufacturer's instructions. The extra reads come from the barcode
  matching, measured on 5 million read pairs per kit (`docs/later/fqtk_vs_cutadapt.md`, 2026-10-03):
  - fqtk assigns 2.72 % more pairs for FlexPrep and 1.91 % for 96-Plex;
  - no pair goes to a different sample;
  - the estimated cost is at most a few wrongly assigned pairs per million.

The remaining 28 % difference with zealbc1 stays open; zealbc1 is not the truth, and accuracy will be measured on
simulated data.
