---
title: "Marker union rarefaction"
subtitle: "How the union of donor sites grows with donors: a demographic model of teosinte, the simulations that would complete it, and the curves fitted to the four-donor chr10 test (2026-10-07)"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

The marker union is every site found in at least one donor (math supplement, Section 5.1). Its size sets the cost of the
union counts, of gap filling and of every per-line table, so it is projected before a run over all donors. This notebook
(i) states what is known about the donors' populations and the union it implies, (ii) writes down the simulations that would
turn that into a predicted curve, (iii) fits empirical curves to the data we have, and (iv) reads the fits against the
demographic picture, including what to expect from other taxa. The simulations are not run: they are written here to be run
once the union over all samples exists. The math supplement's Section 5.6 summarises sections 3-4.

::: {.callout-note}
## In short

- **Demography (Chen et al. 2022).** Mexicana's effective size fell about threefold between ~30 and ~5 thousand years ago
  (MSMC2), and mexicana accessions carry about **6x** the private variation a neutral equilibrium predicts. Mexicana races are
  topographically isolated populations (Sánchez-González et al. 2018). Expected: real alleles saturate within a population,
  each new population adds a block of private alleles, and false alleles add a constant number per donor.
- **Simulations proposed** (section 2): the expected site frequency spectrum of the donors under (S1) mexicana's Ne history,
  (S2) races as demes, (S3) the Zea species tree with ZEAL's donor composition; Eq. 1 turns each into a union curve.
- **Empirical fits** (four mexicana donors, two races): Michaelis-Menten ($U_\infty \approx 414$ thousand sites on chr10),
  Watterson, and Watterson with false alleles ($\varepsilon_f \approx 13$ thousand per donor): 263, 253 and 304 thousand sites at 10
  donors. The false-allele term's linear growth is what sets the last apart (section 3.7).
- **Reading.** Chen's numbers are genome-wide; scaled to our low-copy ranges (section 4), most of the private term
  $\varepsilon_f$ is false alleles, as the QC-set precision implies; the discovery sensitivity is not determined from Chen.
  Other taxa: Zea as a whole carries 2.3x mexicana's SNPs in Chen's panel, an upper bound for the union's growth when ZEAL's
  parviglumis and other donors are added.
:::

```{r setup}
library(data.table)
library(ggplot2)
theme_set(theme_bw(base_size = 14) + theme(legend.text = element_text(size = 12)))
```

## 1. A demographic model of the donors

### 1.1 Who the donors are

ZEAL has 96 teosinte donors (`meta/samples.csv`). Mexicana races from the passport
table (`meta/accessions.csv`, J2Teo); teosinte races are "geographical populations spatially isolated by the topography in
Mexico and Central America" (Sánchez-González et al. 2018).

```{r donors-table}
knitr::kable(data.table(
  taxon = c("mexicana", "parviglumis", "diploperennis", "luxurians", "huehuetenangensis"),
  donors = c(43, 40, 6, 5, 2),
  `mexicana donors by race` = c("Mesa Central 21, Chalco 14, Durango 5, Nobogame 3", "", "", "", "")
))
```

### 1.2 What Chen et al. (2022) measured

Chen et al. resequenced 237 teosinte accessions (22x) with 507 maize lines. Their Supplementary Table 2 (diversity map),
Supplementary Table 4 (BPP demography), Fig. 2a (F~ST~) and Supplementary Fig. 3 (MSMC2 Ne through time) give:

```{r chen-diversity}
chen <- data.table(
  taxon = c("mexicana", "parviglumis", "diploperennis", "perennis", "luxurians", "nicaraguensis", "huehuetenangensis"),
  accessions = c(81, 70, 20, 19, 14, 14, 5),
  snps_M = c(31.32, 41.51, 11.66, 15.29, 9.04, 7.41, 7.99),
  taxon_specific_M = c(4.26, 5.97, 1.13, 1.70, 0.48, 0.37, 0.64),
  specific_downsampled_M = c(1.82, 1.14, 1.06, 1.70, 0.58, 0.41, 1.00),
  pi = c(0.00804, 0.00847, 0.00629, 0.01038, 0.00592, 0.00543, 0.00572),
  fst_mexicana = c(NA, 0.068, 0.277, 0.277, 0.290, 0.270, 0.248),
  fst_maize = c(0.158, 0.101, 0.356, 0.354, 0.364, 0.342, 0.330)
)
knitr::kable(chen, digits = 5, col.names = c("taxon", "accessions", "SNPs (M)", "taxon-specific SNPs (M)",
                                              "taxon-specific, downsampled (M)", "π per bp", "F_ST with mexicana",
                                              "F_ST with maize"))
```

F~ST~ with maize is Chen's cultivated panel (507 inbreds; B73's subspecies): parviglumis and mexicana are closest to maize
(0.101, 0.158), the other taxa 0.33-0.36, so B73 is the outgroup of Eq. 1 for every donor, nearer to some taxa than others.
All Zea together (744 genomes, maize included): 70.73 M SNPs, 16.90 M taxon-specific. Mexicana's chr10 windows (Fig. 2b data):
mean $\pi$ = 0.00903. SNPs specific to one accession (Fig. 2d data): mexicana median 402,200 per accession (166,559-518,710),
parviglumis 322,355.

**Ne through time** (Supplementary Fig. 3, MSMC2; values read off the figure, approximate):

| Time before present | mexicana | parviglumis |
|---|---|---|
| ~30 ky | ~82,000 | ~86,000 |
| ~20 ky | ~77,000 | ~86,000 |
| ~12 ky | ~60,000 | ~86,000 |
| ~8 ky | ~44,000 | ~76,000 |
| ~5 ky to now | ~27,000 | ~62,000, then ~49,000 since ~4 ky |

The perennial and Central American taxa (diploperennis, luxurians, nicaraguensis, huehuetenangensis, perennis) decline from
35-60 thousand around 50-100 ky to 10-25 thousand around 8-10 ky, then rise again (diploperennis and luxurians to ~42-47
thousand). BPP divergence times (Supplementary Table 4, years): the (maize, parviglumis, mexicana) clade 30,467; with
huehuetenangensis 67,800; with diploperennis and perennis 121,533; the Zea crown 124,233; Zea-Tripsacum 653,733. BPP's Ne for
mexicana is 171,508 (95 % HPD 146,817-197,008), on a different scale from MSMC2's; only the shape of the MSMC2 history is used.

### 1.3 The union under a coalescent

Take B73 as the outgroup: a union site is a position where at least one donor carries an allele different from B73. Under
infinite sites, such alleles arise (a) on the B73 lineage or on the donors' lineage before their common ancestor, carried by
every donor (the fixed differences, $C$ of them), or (b) on the donors' genealogy, carried by $k$ of the $n$ donors. Each donor
finds a carried allele with probability $\eta$ (discovery sensitivity), and calls $\varepsilon_f$ false alleles of its own. Then

$$
\mathbb{E}[U_n] = C\,\big[1 - (1 - \eta)^n\big] \;+\; \theta \sum_{k=1}^{n-1} \xi_k(n)\, \big[1 - (1 - \eta)^k\big] \;+\; \varepsilon_f\, n
\tag{1}
$$

with $\theta = 4 N_e \mu L$ over the sites considered and $\xi_k(n)$ the expected site frequency spectrum of $n$ donors in units of
$\theta$ (the expected length of the genealogy's branches that subtend exactly $k$ donors, scaled). **Demography and structure
enter only through $\xi_k(n)$.** For one population at constant size (Wright-Fisher equilibrium), $\xi_k(n) = 1/k$; with
$\eta = 1$ and $\varepsilon_f = 0$, Eq. 1 is Watterson's (1975)

$$
\mathbb{E}[U_n] = C + \theta \sum_{k=1}^{n-1} \frac{1}{k} = C + \theta H_{n-1}
\tag{2}
$$

How $\xi_k(n)$ departs from $1/k$:

- **A recent contraction** (mexicana, threefold over the last ~30 ky, recent relative to the current $2 N_e \approx 54$ ky):
  fewer singletons and low-frequency alleles than $1/k$, more at intermediate frequencies; the union saturates faster than
  $H_{n-1}$.
- **Structure** (isolated races, low migration): donors from one deme share recent ancestry; a donor from a new deme adds the
  long branches of that deme, a block of private alleles. The union grows by steps as demes are added.
- **Several taxa**: the same, at the scale of the species tree (Section 5).

### 1.4 The equilibrium baseline

zealgt-old notebook 01 (2026-09-26) used the large-population version of Eq. 2: site frequencies $p$ in the population with
the neutral spectrum $g(p) \propto 1/p$ (a flat spectrum as the contrast), a carried allele found with probability $\eta$,
no fixed differences and no false alleles:

$$
\mathbb{E}[U_n] = S \sum_{p} g(p) \left[ 1 - (1 - \eta p)^n \right], \qquad
\hat U_n = U_2^{\mathrm{obs}} \, \frac{\mathbb{E}[U_n]}{\mathbb{E}[U_2]}
\tag{3}
$$

```{r data}
donors <- data.table(donor = c("Zx.0540_P3", "Zx.0570_P2", "Zx.0550_P4", "Zx.0580_P2"),
                     race = c("Durango", "Nobogame", "Durango", "Nobogame"),
                     own_sites = c(32750, 48251, 85952, 82601))
unions <- data.table(set = c("pilot pair", "+ Zx.0550_P4", "+ Zx.0580_P2", "all four"),
                     donors = c(2, 3, 3, 4), members = list(1:2, 1:3, c(1, 2, 4), 1:4),
                     sites = c(70210, 128964, 125970, 170729))
pat <- fread("union_rarefaction/patterns.tsv", header = FALSE, col.names = c("pattern", "n"),
             colClasses = c("character", "numeric"))
z <- do.call(rbind, lapply(strsplit(pat$pattern, ""), as.integer))
```

```{r afs, fig.height = 5.5, fig.width = 11}
g <- (1:2000) / 2000
g1 <- head(g, -1)
spectra <- list(`neutral 1/p` = (1 / g1) / sum(1 / g1), flat = rep(1 / length(g1), length(g1)))
afs <- rbindlist(lapply(names(spectra), function(k) rbindlist(lapply(c(0.05, 0.1, 0.15, 0.3), function(eta) {
  eu <- sapply(1:10, function(n) sum(spectra[[k]] * (1 - (1 - eta * g1)^n)))
  data.table(spectrum = k, eta = sprintf("η = %.2f", eta), donors = 2:10, sites = unions$sites[1] * eu[2:10] / eu[2])
}))))
afs[, spectrum := factor(spectrum, levels = names(spectra))]
lab <- unions[, .(donors, sites, label = paste0(set, ": ", format(sites, big.mark = ",")), vj = c(1.6, -0.6, 1.6, 1.6))]
plot_curves <- function(curves, title, colour) {
  ggplot(curves, aes(donors, sites / 1e3, colour = eta)) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 1.5) +
    geom_point(data = lab, aes(donors, sites / 1e3), inherit.aes = FALSE, size = 4, shape = 21, fill = "black") +
    geom_text(data = lab, aes(donors, sites / 1e3, label = label, vjust = vj), inherit.aes = FALSE, hjust = -0.08,
              size = 3.5) +
    scale_x_continuous(breaks = 1:10) +
    labs(title = title, x = "Donors in the union", y = "Union sites (thousands)", colour = colour)
}
plot_curves(afs, "Equilibrium baseline (Eq. 3) and the observed unions", "Discovery sensitivity") + facet_wrap(~spectrum)
```

At 4 donors Eq. 3 gives
`r paste(round(afs[spectrum == "neutral 1/p" & donors == 4, range(sites)] / unions$sites[1], 2), collapse = "-")`x
(neutral) the 2-donor union; the test gave `r round(unions$sites[4] / unions$sites[1], 2)`x. The equilibrium baseline has no
private term: neither the false alleles nor the demes of section 1.3.

## 2. Proposed simulations (not run)

Each simulation produces $\xi_k(n)$ for the donor sets of interest; Eq. 1 turns it into a union curve with $\theta$ from
Chen's $\pi$, $\eta$ and $\varepsilon_f$ from our data (section 4), and $C$ from the fixed differences (or fitted). Tool: msprime
with a demes model (demography as a YAML file), infinite sites, no recombination needed for expectations (or chr10's
recombination map to get the variance along the chromosome); about $10^4$ replicate genealogies per donor set.

| | Model | Parameters (source) | Donor sampling | Output |
|---|---|---|---|---|
| **S0** | one population, constant $N_e$ | none | $n$ haplotypes | $\xi_k(n) = 1/k$: checks the pipeline against Eq. 2 |
| **S1** | mexicana, one population, piecewise $N_e(t)$ | Supplementary Fig. 3 steps (section 1.2; digitise the figure for exact values); generation time 1 year | $n = 1, \dots, 43$ | the contraction's effect on $\xi_k(n)$ and the union's saturation |
| **S2** | mexicana races as demes (Mesa Central, Chalco, Durango, Nobogame), each with S1's history, symmetric migration $m$ | $m$ swept so that within-race vs between-race diversity matches Chen's accession-specific SNPs (6x neutral, section 4) and mexicana's K = 2 | ZEAL's mexicana donors: 21 / 14 / 5 / 3 by race, added in random order and race by race | the step a new race adds; the 43-donor mexicana union |
| **S3** | the Zea species tree | BPP divergence times and Ne (Supplementary Table 4), MSMC2 histories per taxon, introgression ignored at first | ZEAL's 96 donors (43 / 40 / 6 / 5 / 2 by taxon), in random order and taxon by taxon | the union over all donors, and the block each taxon adds |

Calibration targets for S1-S3 (Chen's panel, 22x individual genomes): per-taxon SNP counts, taxon-specific and
downsampled taxon-specific SNPs, $\pi$, F~ST~ (section 1.2). Chen's samples are not ZEAL's donors (none of the six donors'
collection numbers is among them), so the simulations use Chen for the populations and ZEAL only for the sampling. The
targets should be Chen's counts **within our low-copy ranges** (their SNP positions intersected with `lowcopy_chr1-10.bed`),
not the genome-wide totals of section 1.2: the share of their SNPs in our ranges is unknown (section 4).

## 3. Empirical curves fitted to the data

### 3.1 The data

Four mexicana donors from two races, chr10, tier-A sites (step 4 of variant discovery). Own sites from the tier-A VCFs; unions
from `BIALLELIC_UNION` (2 and 4 donors: GENOTYPE runs, jobs 1111642 and 1134902) and from the same bcftools merge, norm and
biallelic filter outside the pipeline (3 donors: jobs 1143221, 1143222). The **sharing pattern** of a union site is which
donors' own sites include it, $z \in \{0, 1\}^4$ (job 1143239, `union_rarefaction/patterns.tsv`). The Durango donors were
collected about 2 km apart (Nombre de Dios, Durango), the Nobogame donors about 3 km apart (Guadalupe y Calvo, Chihuahua).

```{r data-tables}
knitr::kable(donors, format.args = list(big.mark = ","), col.names = c("donor", "race", "own tier-A sites"))
knitr::kable(unions[, .(set, donors, sites)], format.args = list(big.mark = ","),
             col.names = c("union", "donors", "union sites"))
knitr::kable(pat[order(-n)], format.args = list(big.mark = ","), col.names = c("pattern z", "union sites"))
```

### 3.2 The rarefaction curve

As in species-richness rarefaction, donors are the samples and sites the species. The rarefaction curve is the mean union over
every set of $n$ donors, computed exactly from the sharing patterns:

$$
\bar U_n = \binom{4}{n}^{-1} \sum_{|D| = n} \; \sum_{z} N_z \, \mathbb{1}\Big[\max_{d \in D} z_d = 1\Big], \qquad n = 1, \dots, 4
\tag{4}
$$

```{r rarefaction}
subsets <- unlist(lapply(1:4, function(n) combn(4, n, simplify = FALSE)), recursive = FALSE)
sub_dt <- rbindlist(lapply(subsets, function(d) {
  data.table(donors = length(d), set = paste(donors$donor[d], collapse = " + "),
             sites = sum(pat$n[apply(z[, d, drop = FALSE], 1, max) == 1]))
}))
rare <- sub_dt[, .(sites = mean(sites)), by = donors]
knitr::kable(rare, format.args = list(big.mark = ","), col.names = c("donors n", "rarefied union (Eq. 4)"))
```

```{r curve-fits}
# harmonic number H_n, continuous in n: digamma(n + 1) + Euler's gamma
harm <- function(n) digamma(n + 1) - digamma(1)
models <- list(
  `Michaelis-Menten` = list(f = sites ~ um * donors / (k + donors), start = list(um = 3e5, k = 4)),
  `Watterson + false alleles` = list(f = sites ~ cc + th * harm(donors) + eps * donors,
                                     start = list(cc = 1e4, th = 5e4, eps = 1e4)),
  Watterson = list(f = sites ~ cc + th * harm(donors), start = list(cc = 1e4, th = 5e4)),
  Logarithmic = list(f = sites ~ a + b * log(donors), start = list(a = 5e4, b = 5e4)),
  Power = list(f = sites ~ a * donors^b, start = list(a = 6e4, b = 0.7)),
  `Square root` = list(f = sites ~ a * sqrt(donors), start = list(a = 6e4)),
  `Negative exponential` = list(f = sites ~ um * (1 - exp(-k * donors)), start = list(um = 3e5, k = 0.2))
)
sym <- c(cc = "c", th = "θ_η", eps = "ε_f", um = "U_∞", k = "K", a = "a", b = "b")
emp <- lapply(models, function(m) nls(m$f, data = rare, start = m$start, control = nls.control(maxiter = 500)))
emp_table <- rbindlist(lapply(names(emp), function(k) {
  cf <- coef(emp[[k]])
  data.table(model = k, parameters = paste(sprintf("%s = %s", sym[names(cf)], signif(cf, 3)), collapse = ", "),
             rmse = sqrt(mean(residuals(emp[[k]])^2)),
             at_10 = predict(emp[[k]], data.table(donors = 10)))
}))
plot_fits <- function(keep, title) {
  cv <- rbindlist(lapply(keep, function(k) {
    data.table(model = k, donors = seq(1, 10, 0.1), sites = predict(emp[[k]], data.table(donors = seq(1, 10, 0.1))))
  }))
  ggplot(cv, aes(donors, sites / 1e3, colour = model)) +
    geom_line(linewidth = 0.9) +
    geom_point(data = sub_dt, aes(donors, sites / 1e3), inherit.aes = FALSE, colour = "grey50", size = 2) +
    geom_point(data = rare, aes(donors, sites / 1e3), inherit.aes = FALSE, size = 4, shape = 21, fill = "black") +
    scale_x_continuous(breaks = 1:10) +
    labs(title = title, subtitle = "black: rarefied union (Eq. 4); grey: each of the 15 donor sets",
         x = "Donors in the union", y = "Union sites (thousands)", colour = NULL)
}
mm <- coef(emp$`Michaelis-Menten`)
wf <- coef(emp$`Watterson + false alleles`)
```

### 3.3 Michaelis-Menten

Species accumulation curves are routinely fitted with (supplement Eq. 21; Soberón and Llorente 1993)

$$
\bar U_n = \frac{U_\infty\, n}{K + n}
\tag{5}
$$

the asymptote $U_\infty$ and the number of donors $K$ that reach half of it. Fitted: $U_\infty$ =
`r format(signif(mm[["um"]], 3), big.mark = ",")`, $K$ = `r round(mm[["k"]], 1)`, RMSE
`r round(emp_table[model == "Michaelis-Menten", rmse])` sites. If each site is found by each donor with rate $k_s$ and $k_s$ is
gamma-distributed across sites (shape $\alpha$, scale $1/K$) (supplement Eq. 22),

$$
\bar U_n = U_\infty \left[ 1 - \left( 1 + \frac{n}{K} \right)^{-\alpha} \right]
\tag{6}
$$

which is Eq. 5 at $\alpha = 1$, the negative exponential as $\alpha \to \infty$, logarithmic growth (Fisher's log-series) as
$\alpha \to 0$ and power growth for $\alpha < 0$. Since $k_s \approx \eta p_s$, $\alpha$ is the shape of the spectrum near
$p = 0$: the neutral $1/p$ is $\alpha \to 0$; Michaelis-Menten is a spectrum with fewer rare alleles than neutral and a finite
pool of sites.

### 3.4 Watterson

Eq. 1 at equilibrium, $\xi_k(n) = 1/k$, with $\eta$ absorbed into the fitted constants and no private term: the union grows with
the harmonic number, like $\ln n$, and has no ceiling.

$$
\bar U_n = c + \theta_\eta H_n
\tag{7}
$$

```{r watterson}
wa <- coef(emp$Watterson)
wa_res <- round(residuals(emp$Watterson))
```

Fitted: $c$ = `r format(round(wa[["cc"]]), big.mark = ",")`, $\theta_\eta$ = `r format(round(wa[["th"]]), big.mark = ",")`, RMSE
`r format(round(emp_table[model == "Watterson", rmse]), big.mark = ",")` sites. $c$ is negative, which Eq. 2 does not allow
(fixed differences cannot be fewer than none), and the residuals follow a pattern, not noise (`r paste(format(wa_res, big.mark = ","), collapse = ", ")`
for $n = 1, \dots, 4$: above, below, below, above): the curve bends too early.

### 3.5 Watterson with false alleles

Eq. 7 plus the private term of Eq. 1 (supplement Eq. 23):

$$
\bar U_n = c + \theta_\eta H_n + \varepsilon_f\, n
\tag{8}
$$

```{r watterson-false}
wf_cor <- cov2cor(vcov(emp$`Watterson + false alleles`))
wf_se <- summary(emp$`Watterson + false alleles`)$coefficients[, 2]
```

Fitted: $c$ = `r format(round(wf[["cc"]]), big.mark = ",")`, $\theta_\eta$ = `r format(round(wf[["th"]]), big.mark = ",")`,
$\varepsilon_f$ = `r format(round(wf[["eps"]]), big.mark = ",")` per donor, RMSE
`r format(round(emp_table[model == "Watterson + false alleles", rmse]), big.mark = ",")` sites. $c$ is still negative; three
parameters on four points leave one degree of freedom, so this is a consistency check.

### 3.6 The three curves together

```{r main-plot}
main_models <- c("Michaelis-Menten", "Watterson", "Watterson + false alleles")
plot_fits(main_models, "Rarefaction curve and the three fitted curves")
```

```{r main-table}
knitr::kable(emp_table[model %in% main_models], digits = 0, format.args = list(big.mark = ","),
             col.names = c("model", "fitted parameters", "RMSE (sites)", "predicted at 10 donors"))
proj <- data.table(donors = c(4, 6, 10, 20, 43))
proj[, `:=`(mm = predict(emp$`Michaelis-Menten`, proj), watterson = predict(emp$Watterson, proj),
            wf = predict(emp$`Watterson + false alleles`, proj),
            wf_real = wf[["cc"]] + wf[["th"]] * harm(donors), wf_false = wf[["eps"]] * donors)]
knitr::kable(proj, digits = 0, format.args = list(big.mark = ","),
             col.names = c("donors", "Michaelis-Menten", "Watterson", "Watterson + false alleles",
                           "of which c + θ_η H_n", "of which ε_f n"))
```

### 3.7 Why Watterson with false alleles diverges from the other curves

The curves agree on the four points and part beyond them, and they part into two groups: those whose growth fades
(logarithmic, negative exponential, Watterson, Michaelis-Menten: 237-263 thousand sites at 10 donors, Appendix A) and those
with a term that never fades (Watterson with false alleles and the power curve: 304-331 thousand).

1. **The data grow faster than a logarithm and slower than a line.** The free power curve puts the exponent at 0.71. Watterson
   (Eq. 7) can only grow like $\ln n$; its residuals run above, below, below, above (section 3.4), so it bends too early and falls short beyond the
   data. That is also why it needs a negative $c$.
2. **The $\varepsilon_f n$ term takes the excess growth.** With it, $\theta_\eta$ drops from
   `r format(round(wa[["th"]]), big.mark = ",")` to `r format(round(wf[["th"]]), big.mark = ",")`, $\varepsilon_f$ takes
   `r format(round(wf[["eps"]]), big.mark = ",")` per donor, and the RMSE falls from
   `r format(round(emp_table[model == "Watterson", rmse]), big.mark = ",")` to
   `r format(round(emp_table[model == "Watterson + false alleles", rmse]), big.mark = ",")` sites. Beyond the data the linear
   term keeps adding $\varepsilon_f$ per donor while $\theta_\eta H_n$ flattens; past about 15 donors $\varepsilon_f n$ is the larger
   part (table above).
3. **The split between $\theta_\eta$ and $\varepsilon_f$ is poorly determined.** Their estimates are correlated at
   `r round(wf_cor["th", "eps"], 2)` (standard errors `r format(round(wf_se[["th"]]), big.mark = ",")` and
   `r format(round(wf_se[["eps"]]), big.mark = ",")`): many combinations fit the four points almost equally and differ widely at
   10 donors and beyond, so the projection is less certain than these standard errors suggest.

**Where Michaelis-Menten puts the false alleles.** Eq. 5 has no private term, so a constant $\varepsilon_f$ per donor is absorbed
by both parameters: the only way it can imitate growth that never fades is to saturate later and higher, raising $K$ and
$U_\infty$ (for $K \gg n$, Eq. 5 is nearly the line $(U_\infty / K)\, n$). Fitting Eq. 5 to the full rarefied curve and to its
real-allele part only (Eq. 8 without $\varepsilon_f n$) shows the size of the effect:

```{r mm-split}
mm_real <- nls(sites ~ um * donors / (k + donors), data = rare[, .(donors, sites = wf[["cc"]] + wf[["th"]] * harm(donors))],
               start = list(um = 3e5, k = 4))
mm_split <- rbindlist(lapply(list(`full curve (real + false)` = emp$`Michaelis-Menten`, `real-allele part only` = mm_real),
  function(m) data.table(um = coef(m)[["um"]], k = coef(m)[["k"]], slope = coef(m)[["um"]] / coef(m)[["k"]])),
  idcol = "fitted to")
knitr::kable(mm_split, digits = c(0, 0, 1, 0), format.args = list(big.mark = ","),
             col.names = c("Eq. 5 fitted to", "U_∞", "K", "initial gain U_∞ / K"))
```

The private term nearly doubles the asymptote and delays saturation. So $U_\infty$ is not a ceiling on real alleles, and once
donors keep adding false or population-private sites at a constant rate Eq. 5 will fall short. The real-allele part comes from
Eq. 8's split of $\theta_\eta$ and $\varepsilon_f$, which is poorly determined (point 3), so its $U_\infty$ is an illustration, not
an estimate.

The four donors cannot tell the two groups apart; the next donors can, since the groups already differ at 6 donors
(table above). Donors from new races (the six-donor run) test whether the private term is real: section 4 reads it as about a
third population-private variation and two thirds false alleles.

## 4. Interpretation in the light of the demographic model

Our counts are all within the low-copy ranges (`lowcopy_chr1-10.bed`: 367 Mb, about 17 % of the 2.2 Gb B73 genome, roughly
its unmasked part; chr10 26.1 Mb). Chen's counts are genome-wide. The share of **Chen's SNPs** that falls in our ranges is not
known: about 17 % if SNPs were spread evenly per base, more if Chen's filters on maize-mapped reads kept mostly SNPs in unique
sequence, up to nearly all of them. Every Chen number below is given at both bounds; intersecting Chen's SNP positions with our
BED would settle it (their VCF is not in `agent/papers/chen2022/`).

```{r chen-numbers}
bounds <- c(`17 % (even per base)` = 0.17, `~100 % (unique sequence)` = 1)
chr10_share <- 26.1 / 367
a_hap <- sum(1 / (1:161))
own <- range(donors$own_sites)
til18_diffs <- 444982
theta_pi <- 0.00903 * 26.1e6
eta_lo <- own[1] / til18_diffs
eta_hi <- wf[["th"]] / theta_pi
singletons_neutral_acc <- 2 * (31316555 / a_hap) / 162
per_donor <- rbindlist(lapply(names(bounds), function(b) {
  private_real <- 402200 / 2 * bounds[[b]] * chr10_share * c(eta_lo, eta_hi)
  data.table(share = b, theta_counts = 31316555 * bounds[[b]] * chr10_share / a_hap,
             real_private = paste(format(round(private_real, -1), big.mark = ","), collapse = "-"),
             false_share = paste(round(100 * (1 - rev(private_real) / wf[["eps"]])), collapse = "-"))
}))
```

1. **Discovery sensitivity $\eta$ is not determined from Chen.** Routes to how much variation a mexicana haplotype carries in
   chr10's ranges:
   - Chen's chr10 $\pi$ (0.00903) times 26.1 Mb: $\theta \approx$ `r format(signif(theta_pi, 3), big.mark = ",")`, so
     $\eta = \theta_\eta / \theta \approx$ `r round(eta_hi, 2)`; this assumes Chen's per-base $\pi$ holds for every base of our
     ranges.
   - Chen's 31.3 M mexicana SNPs in our ranges, by Watterson ($H_{161} \approx$ `r round(a_hap, 2)`): $\theta \approx$
     `r paste(format(round(per_donor$theta_counts, -3), big.mark = ","), collapse = " to ")` at the two bounds, so
     $\eta \approx$ `r paste(round(wf[["th"]] / per_donor$theta_counts, 2), collapse = " to ")`; above 1 at the 17 % bound,
     which is impossible.
   - The TIL18 assembly differs from B73 at `r format(til18_diffs, big.mark = ",")` sites in chr10's ranges (AnchorWave), more
     than Chen's whole mexicana panel there at the 17 % bound: our donors' own sites give
     $\eta \approx$ `r sprintf("%.2f-%.2f", eta_lo, own[2] / til18_diffs)` (zealgt-old's ~0.1). This is the most direct route
     for our discovery; $\pi$'s `r round(eta_hi, 2)` is an upper value.
2. **The excess of private variation is genome-wide.** Under Eq. 2 the expected number of singletons equals $\theta$: Chen's
   31.3 M mexicana SNPs over 161 haplotype intervals predict about `r format(signif(singletons_neutral_acc, 2), big.mark = ",")`
   per diploid accession, and Chen observe 402,200 accession-specific SNPs, about
   `r round(402200 / singletons_neutral_acc)`x more. A contraction would give fewer; isolated populations, mapping errors and
   private insertions in repeats give more. Until Chen's SNPs are placed in our ranges, this excess is not evidence that
   population-private blocks dominate our union.
3. **Most of the private term $\varepsilon_f$ is false alleles.** $\varepsilon_f$ counts the sites per donor that no other
   donor shares (`r format(round(wf[["eps"]]), big.mark = ",")`), not the donor's own sites (about 62 thousand, rarefied
   $n = 1$). Real private sites found per donor: half of 402,200 per haplotype, times the share in our ranges, times chr10's
   share (26.1 / 367), times $\eta$ = `r sprintf("%.2f-%.2f", eta_lo, eta_hi)`:

   ```{r eps-split}
   knitr::kable(per_donor[, .(share, real_private, false_share)],
                col.names = c("Chen's SNPs in our ranges", "real private sites per donor", "share of ε_f that is false (%)"))
   ```

   This agrees with the QC set: tier-A precision 0.67-0.74 makes a quarter to a third of a donor's ~62 thousand own sites false,
   16-21 thousand per donor. False alleles are mostly private, so they fill $\varepsilon_f$; real alleles are mostly shared and
   go into $c + \theta_\eta H_n$. "Most of $\varepsilon_f$ is false" and "a third of own sites is false" are the same fact
   seen from the union and from the donor.
4. **Which curve.** Within one population the contraction predicts saturation of real alleles: the Michaelis-Menten shape. What
   keeps the union growing in Eq. 8 is $\varepsilon_f n$, mostly false alleles, a property of our discovery, not of teosinte.
   So the gap between the two groups of curves (section 3.7) is mainly a question of how many false alleles each new donor adds.
   The six-donor run (Mesa Central and Chalco donors, `docs/runs/genotype_union_6donors_races_chr10.md`) measures the step a new
   race adds in our ranges, which Chen's genome-wide numbers cannot give.

## 5. What to expect from the other taxa

ZEAL's donors are 43 mexicana, 40 parviglumis and 13 from three other taxa. In Chen's panel:

- **Parviglumis is close to mexicana** (F~ST~ 0.068, the lowest between teosinte taxa) but more diverse (41.5 M SNPs against
  31.3 M; $\pi$ 0.0085 against 0.0080) and with a smaller, later decline (MSMC2: 86 to 49 thousand). It shares much of mexicana's
  variation, but its 5.97 M taxon-specific SNPs (14 % of its SNPs) are a block the mexicana union does not contain, and its larger
  recent Ne means more low-frequency variation per donor.
- **ZEAL's three other taxa** (diploperennis 6 donors, luxurians 5, huehuetenangensis 2) are distant (F~ST~ with mexicana
  0.25-0.29) and carry 1.13, 0.48 and 0.64 M taxon-specific SNPs; with 2-6 donors each, they add their blocks almost entirely
  with their first donors.
- **An upper bound.** All Zea together carry 70.7 M SNPs against mexicana's 31.3 M (2.3x; maize included). The union of all ZEAL
  donors cannot grow by more than that factor over the mexicana union at the same depth of sampling, and will grow by less,
  since ZEAL has few donors of the distant taxa and Chen's maize-specific SNPs (1.5 M and 0.9 M, tropical and temperate) do not
  enter.

These ratios are genome-wide; they carry over to our ranges only if every taxon has the same share of its SNPs there, and
taxon-specific SNPs may sit more often in repeats, so they likely overstate what other taxa add in our ranges.
A first-order projection: the mexicana union at 43 donors times a taxon factor between ~1.2 (parviglumis-specific SNPs only)
and 2.3 (all Zea). Simulation S3 replaces that range with a curve; the union over all samples measures it.

## Caveats

- Four donors, two races, all mexicana: every curve in section 3 fits them; the choice between curves rests on section 4, not
  on the fit.
- The rarefaction mean averages over donors; a donor with few lines or low coverage finds fewer sites and adds fewer.
- Section 4's numbers mix Chen's discovery (individual genomes, 22x, strict filters) with ours (BC1 pools); $\eta$ and the
  $\varepsilon_f$ split are orders of magnitude.

## Appendix A: other accumulation curves, set aside

All fitted to the same four points (Eq. 4); the table and plot include the three curves of section 3.

| Model | $\bar U_n =$ | Spectrum it corresponds to | Why set aside |
|---|---|---|---|
| Logarithmic | $a + b \ln n$ | neutral, large $n$; Eq. 6 at $\alpha \to 0$ | the same as Watterson |
| Power | $a\, n^{b}$ | $p^{a' - 1}$, $-1 < a' < 0$, $b = -a'$; Eq. 6 at $\alpha < 0$ | $b = 0.71$ needs $g(p) \propto p^{-1.71}$, rapid population growth, which Chen's MSMC2 history contradicts |
| Square root | $a \sqrt{n}$ | power with $b = 1/2$ | fixes the 4-to-1 donor ratio at 2 against 2.75 observed; worst fit |
| Negative exponential | $U_\infty (1 - e^{-kn})$ | every site at one frequency; Eq. 6 at $\alpha \to \infty$ | no spread of frequencies |

```{r appendix-a}
knitr::kable(emp_table, digits = 0, format.args = list(big.mark = ","),
             col.names = c("model", "fitted parameters", "RMSE (sites)", "predicted at 10 donors"))
plot_fits(names(emp), "Every accumulation curve fitted")
```

## Appendix B: pattern-level models, set aside

These fit the 15 sharing patterns (170,729 sites) instead of the four rarefied points, with a discovery sensitivity per donor.
They are set aside because they have no false-allele or population-private class: the many private patterns (`0010`, `0001`)
can only come from rare alleles and high $r_d$, so their parameters absorb errors and structure. They are the starting point
for fitting S1-S3's spectra to the patterns (a likelihood that allows AIC and likelihood-ratio tests, and per-donor
predictions), once a private class is added.

### Version 2: a sensitivity per donor

$$
g_2(p) = (1 - \phi)\, \frac{p^{a - 1}}{\sum_{q} q^{a - 1}} + \phi\, \mathbb{1}[p = 1], \qquad
P(z) = \sum_{p} g_2(p) \prod_{d = 1}^{4} (r_d\, p)^{z_d} (1 - r_d\, p)^{1 - z_d}
\tag{B1}
$$

$$
\log L = \sum_{z \neq 0000} N_z \log \frac{P(z)}{1 - P(0000)}, \qquad \hat S = U_{1234}^{\mathrm{obs}} / (1 - P(0000))
\tag{B2}
$$

### Version 3: a shared class

A fraction $\phi$ of sites that every donor carries and finds with the same probability $e$ (the fixed differences $C$ of
Eq. 1); the rest as B1 without the point mass ($g_3(p) \propto p^{a - 1}$):

$$
P(z) = (1 - \phi) \sum_{p} g_3(p) \prod_{d = 1}^{4} (r_d\, p)^{z_d} (1 - r_d\, p)^{1 - z_d}
\; + \; \phi \prod_{d = 1}^{4} e^{z_d} (1 - e)^{1 - z_d}
\tag{B3}
$$

```{r pattern-fits}
p_rare <- function(zz, r, w, grid) {
  apply(zz, 1, function(zr) {
    q <- rep(1, length(grid))
    for (d in seq_along(r)) q <- q * if (zr[d] == 1) grid * r[d] else 1 - grid * r[d]
    sum(w * q)
  })
}
spectrum <- function(a) g^(a - 1) / sum(g^(a - 1))
model <- list(
  v2 = list(start = c(rep(-1, 4), -3, log(0.5)),
            par = function(th) list(r = plogis(th[1:4]), phi = plogis(th[5]), a = exp(th[6]), e = NA),
            pz = function(zz, m) p_rare(zz, m$r, c((1 - m$phi) * spectrum(m$a), m$phi), c(g, 1)),
            miss = function(rs, m) {
              sum(c((1 - m$phi) * spectrum(m$a), m$phi) * Reduce(`*`, lapply(rs, function(x) 1 - c(g, 1) * x)))
            }),
  v3 = list(start = c(rep(-1, 4), -3, log(0.5), 1),
            par = function(th) list(r = plogis(th[1:4]), phi = plogis(th[5]), a = exp(th[6]), e = plogis(th[7])),
            pz = function(zz, m) {
              (1 - m$phi) * p_rare(zz, m$r, spectrum(m$a), g) +
                m$phi * apply(zz, 1, function(zr) prod(ifelse(zr == 1, m$e, 1 - m$e)))
            },
            miss = function(rs, m) {
              (1 - m$phi) * sum(spectrum(m$a) * Reduce(`*`, lapply(rs, function(x) 1 - g * x))) +
                m$phi * (1 - m$e)^length(rs)
            })
)
fits <- lapply(model, function(md) {
  nll <- function(th) {
    pz <- md$pz(z, md$par(th))
    -sum(pat$n * log(pz / sum(pz)))
  }
  ft <- optim(md$start, nll, method = "BFGS", control = list(maxit = 2000))
  m <- md$par(ft$par)
  m$S <- sum(pat$n) / (1 - md$miss(m$r, m))
  m$union <- function(rs) m$S * (1 - md$miss(rs, m))
  m$fitted <- sum(pat$n) * md$pz(z, m) / sum(md$pz(z, m))
  m$loglik <- -ft$value
  m
})
par_table <- rbindlist(lapply(names(fits), function(v) {
  m <- fits[[v]]
  data.table(version = sub("v", "", v), r_0540 = m$r[1], r_0570 = m$r[2], r_0550 = m$r[3], r_0580 = m$r[4],
             a = m$a, phi = m$phi, e = m$e, S = round(m$S), log_lik = round(m$loglik, 1))
}))
knitr::kable(par_table, digits = 3, format.args = list(big.mark = ","))
knitr::kable(data.table(pattern = pat$pattern, observed = pat$n, version_2 = round(fits$v2$fitted),
                        version_3 = round(fits$v3$fitted))[order(-observed)],
             format.args = list(big.mark = ","), col.names = c("pattern z", "observed", "Version 2 (B1)", "Version 3 (B3)"))
```

Version 2 goes to the edge of B1 ($\phi \approx 0$, $a \approx 0$) and predicts about half of the sites found by all four
donors (`1111`). Version 3 fits the patterns and the observed unions within 1 %, with a log-likelihood gain of
`r round(fits$v3$loglik - fits$v2$loglik, 1)` for one more parameter.

## References

- Chen L, et al. 2022. Genome sequencing reveals evidence of adaptive variation in the genus *Zea*. *Nat Genet* 54:1736-1745
  (main text and supplement: `agent/papers/chen2022/`).
- Sánchez González JJ, et al. 2018. Ecogeography of teosinte. *PLoS ONE* 13:e0192676.
- Soberón J, Llorente J. 1993. The use of species accumulation functions for the prediction of species richness.
  *Conserv Biol* 7:480-488.
- Watterson GA. 1975. On the number of segregating sites in genetical models without recombination. *Theor Popul Biol*
  7:256-276.
