---
title: "The BC2S3 line pool as a denoiser, and the Zx.0090_P2 family"
subtitle: "What the in-silico pool of each donor's lines does in discovery, and how the six-donor test's failure was narrowed down to two families (chr10, 2026-10-07)"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

Variant discovery runs CRISP per donor on its BC1 pools plus one in-silico pool of all its BC2S3 lines (math supplement,
Section 4), then keeps a site only if that line pool shows an ALT read (the witness veto, Section 4.2), then step 4's tiers
(ALT in at least 2 BC1 pools, likelihood ratio, flags). RTIGER then infers each line's ancestry at the donor's tier-A sites.
In the six-donor union test (`docs/runs/genotype_union_6donors_races_chr10.md`) two of the six donors failed: few sites and
fragmented ancestry. This notebook records what the line pool does for the other four, and how the failure was traced to
the two families rather than to the method or the sequencing.

::: {.callout-note}
## In short

- **The line pool is a denoiser for RTIGER.** Markers from the BC1 pools alone fragment the ancestry of good donors (8-10.5
  donor segments per line on chr10); with the line pool and the veto they give 1-2, like the independent SNP50K calls. Only
  the deepest donor (Zx.0550_P4, 47 thousand markers) stays clean without it.
- **Zx.0090_P2 is a family problem, Zx.0500_P2 a milder one.** Their BC1 pools are normal; their line samples carry a
  background of their own donor's alleles (Zx.0090_P2 4-10 %, Zx.0500_P2 1-3 %, clean lines 0.3-1 %, B73 0.07 %) all along
  chr10, also in lines that SNP50K calls B73, in both sequencing batches. No marker set fixes them.
- **Most likely cause, upstream of sequencing:** a problem in the cross or seed from the wrong ear in the planted packet
  (user). If the line samples are bulks of seedlings, a few off-type seeds of the same donor family in the packet give every
  bulk exactly this kind of background.
:::

```{r setup}
library(data.table)
library(ggplot2)
theme_set(theme_classic(base_size = 12) + theme(strip.background = element_blank()))
pal <- c(REF = "gold", HET = "springgreen4", ALT = "purple4") # nilHMM paint_calls() palette
d <- "line_pool_denoiser"
seg <- fread(file.path(d, "segments_zealgt.tsv"))
snp <- fread(file.path(d, "snp50k_rtiger_calls_lines.tsv"))
cnt <- fread(file.path(d, "private_site_counts.tsv"))
cov <- fread(file.path(d, "coverage.tsv"))
short <- function(x) sub("_P\\d+$", "", x)
```

## The data

Six mexicana donors on chr10: the four of the first test (Zx.0540_P3, Zx.0570_P2 races Durango and Nobogame, Zx.0550_P4,
Zx.0580_P2) and two added for new races, Zx.0500_P2 (Mesa Central) and Zx.0090_P2 (Chalco). GENOTYPE runs: jobs 1134902 and
1146812. Independent reference: zealhmm's RTIGER on the SNP50K panel (zealtiger `fit_rtiger_by_donor.R`: per donor accession,
rigidity 8, ~49 thousand panel markers genome-wide, ~3.4 thousand on chr10), which covers the batch-1 lines of all six donors.

| Donor | Race | BC1 pools | Lines | Line plates |
|---|---|---|---|---|
| Zx.0540_P3 | Durango | 5 | 40 batch-1 | BZea5, BZea6 |
| Zx.0550_P4 | Durango | 4 | 43 batch-1 | BZea5 |
| Zx.0570_P2 | Nobogame | 5 | 44 batch-1 | BZea8, BZea9 |
| Zx.0580_P2 | Nobogame | 5 | 50 batch-1 | BZea7, BZea8 |
| Zx.0500_P2 | Mesa Central | 5 | 11 batch-1 + 11 batch-2 | BZea4; BZeaV2_4 |
| Zx.0090_P2 | Chalco | 5 | 12 batch-1 + 11 batch-2 | BZea3, BZea4; BZeaV2_3 |

## 1. The line pool as a denoiser

### Discovery with and without the line pool

CRISP was also run on each donor's BC1 pools alone (jobs 1148722, 1148870, 1149032), with the pipeline's command and
container. Markers for RTIGER from that run: biallelic SNPs in the low-copy ranges with ALT reads in at least 2 BC1 pools
(step 4's pool rule; no veto, no likelihood or flags); lines counted and RTIGER run as in the pipeline (job 1149090).

| Donor | CRISP, BC1 pools only | CRISP, BC1 pools + line pool | Kept by the veto | Tier A |
|---|---|---|---|---|
| Zx.0540_P3 | 8,099 | 60,312 | 43,925 (73 %) | 32,750 |
| Zx.0570_P2 | 11,741 | 100,331 | 64,640 (64 %) | 48,251 |
| Zx.0550_P4 | 60,634 | 184,394 | 111,740 (61 %) | 85,952 |
| Zx.0580_P2 | 28,226 | 178,789 | 122,413 (68 %) | 82,601 |
| Zx.0500_P2 | 6,903 | 20,518 | 13,749 (67 %) | 9,060 |
| Zx.0090_P2 | 22,104 | 45,606 | 12,477 (27 %) | 7,482 |

(CRISP with the line pool counted after the low-copy clip; jobs 1147356, 1149087.) The line pool multiplies CRISP's calls
3-9 fold, but mostly by adding statistical power: 89 % of the pre-veto calls have ALT reads in at least 2 BC1 pools. Among
those, the veto drops the ones without an ALT read in the lines: removing it would add 17-32 % of sites for the four good
donors and 30 % for Zx.0500_P2, but 204 % for Zx.0090_P2 (upper bounds before step 4's likelihood and flags; job 1149087,
validated against step 4's tables).

### Ancestry from the three marker sets

```{r three-columns, fig.width = 13, fig.height = 9}
z <- seg[, .(source = fifelse(markers == "bc1_only", "CRISP BC1", "CRISP BC1 + BC2S3 pool"), donor = short(donor),
             line, start, end, state)]
s <- snp[chr == 10, .(source = "SNP50K", donor = short(donor), line = name, start = start_bp, end = end_bp, state)]
both <- Reduce(intersect, list(unique(s$line), unique(z[source == "CRISP BC1"]$line),
                               unique(z[source == "CRISP BC1 + BC2S3 pool"]$line)))
dt <- rbind(s, z)[line %in% both]
dt[, row := match(line, rev(sort(unique(line)))), by = donor]
dt[, source := factor(source, levels = c("SNP50K", "CRISP BC1", "CRISP BC1 + BC2S3 pool"))]
dt[, state := factor(state, levels = 0:2, labels = names(pal))]
ggplot(dt) +
  geom_rect(aes(xmin = start / 1e6, xmax = end / 1e6, ymin = row - 0.48, ymax = row + 0.48, fill = state)) +
  scale_fill_manual(values = pal, name = "State") +
  facet_grid(donor ~ source, scales = "free_y", space = "free_y") +
  scale_y_continuous(breaks = NULL, expand = c(0, 0)) + scale_x_continuous(expand = c(0, 0)) +
  labs(title = "chr10 ancestry per line (RTIGER per donor, same lines)", x = "chr10 position (Mb)", y = "lines") +
  theme(strip.text.y = element_text(angle = 0), legend.position = "bottom", panel.spacing.y = unit(4, "pt"))
```

```{r segments-table}
per_line <- dt[, .(segs = sum(state != "REF"), share = sum((end - start) * (as.integer(state) - 1) / 2) / sum(end - start)),
               by = .(donor, source, line)]
tab <- per_line[, .(lines = .N, `donor segments per line (median, lines with any)` = as.numeric(median(segs[segs > 0])),
                    `lines with no introgression (< 2 %)` = sum(share < 0.02)), by = .(donor, source)]
knitr::kable(dcast(tab, donor + lines ~ source, value.var = c("donor segments per line (median, lines with any)",
                                                              "lines with no introgression (< 2 %)")))
```

For the four good donors the line pool removes what the BC1 pools alone call but the lines do not support (false alleles,
pool errors): the right column matches SNP50K, the middle one does not. That is the denoising the veto was meant to do at
RTIGER's input (user: "the BC2S3 pool works as a noise filter for most of the donors"). For the two problem donors neither
marker set gives SNP50K's picture.

## 2. Narrowing down the Zx.0090_P2 problem

Each step below ruled out a candidate.

**1. The calls are broken, the plants are not.** SNP50K RTIGER on the same batch-1 lines shows a normal BC2S3 pattern for both
donors (table above: most lines without introgression on chr10, a few long segments); ours spreads short segments over every
line. RTIGER is fitted per donor in both, so the difference is the markers, not the model.

**2. Not coverage.**

```{r coverage}
cv <- cov[, .(`BC1 pools` = sum(role == "bc1_sample"), `BC1 coverage, summed` = round(sum(mean_coverage[role == "bc1_sample"]), 1),
              lines = sum(role == "nil"), `line coverage, summed` = round(sum(mean_coverage[role == "nil"]), 1)), by = donor]
knitr::kable(cv)
```

The problem donors' BC1 pools are as deep as the others'; they have half as many lines.

**3. Not the BC1 pools.** At each donor's private tier-A sites (sites only that donor called among the six), every BC1 pool is
highest at its own donor's sites; the problem donors' pools look like the others'. BC1-only CRISP finds 6,903 sites for
Zx.0500_P2 against 8,099 for Zx.0540_P3.

```{r pools}
pc <- cnt[role == "bc1_sample", .(pct = 100 * sum(alt_reads) / sum(ref_reads + alt_reads)),
          by = .(sample_donor, own = fifelse(site_set == sample_donor, "own donor's sites", "other donors' sites"))]
knitr::kable(dcast(pc, sample_donor ~ own, value.var = "pct", fun.aggregate = function(x) round(mean(x), 1)),
             col.names = c("pools of", "% non-REF reads at other donors' private sites (mean)",
                           "% at own donor's private sites"))
```

**4. The lines carry a background of their own donor's alleles; B73 does not.**

```{r background, fig.height = 4.5}
lines_snp <- snp[chr == 10, .(chr10_donor = sum((end_bp - start_bp) * state / 2) / sum(end_bp - start_bp)), by = name]
bg <- merge(cnt[role == "nil" & site_set == sample_donor, .(sample, sample_donor, batch, pct = 100 * alt_reads / (ref_reads + alt_reads))],
            lines_snp, by.x = "sample", by.y = "name", all.x = TRUE)
b73 <- cnt[sample_donor %in% c("B73", "B73-bulk") & site_set == "Zx.0090_P2",
           .(group = fifelse(batch == "wgs", "B73 WGS control", fifelse(batch == "bc2s3_batch2", "B73 checks, batch 2",
                                                                       "B73 checks, batch 1")),
             pct = 100 * alt_reads / (ref_reads + alt_reads))]
clean <- bg[!is.na(chr10_donor) & chr10_donor < 0.02]
ggplot(clean, aes(short(sample_donor), pct)) +
  geom_jitter(width = 0.15, height = 0, size = 2, colour = "grey30") +
  scale_y_log10() +
  labs(title = "Lines with no chr10 introgression (SNP50K): % non-REF reads at their own donor's private sites",
       x = NULL, y = "% non-REF reads (log scale)")
knitr::kable(b73[, .(`% non-REF reads at Zx.0090_P2's private sites` = round(mean(pct), 2)), by = group])
```

In lines that SNP50K calls B73 on chr10, the four good donors' lines show the error background (0.3-1 %); Zx.0500_P2's lines
1-3 %, Zx.0090_P2's 4-10 %. B73 shows 0.07 % at the same sites, so these are not paralog or mapping artifacts. The signal also
tracks real chr10 introgression in every donor (correlation with the SNP50K chr10 donor share 0.93-0.99) and not the
genome-wide share (Zx.0090_P2: -0.44): it is real donor DNA on chr10 plus a baseline. A plant can only give 0, ~50 or ~100 %
at a site; a steady few percent in every line means each line sample carries some DNA that is not that plant's.

**5. Not a sequencing batch or a plate.** Zx.0090_P2's batch-1 lines (plate 3) and batch-2 lines (BZeaV2_3) carry the same
background; they were sampled and sequenced separately. Most batch-1 B73 checks are clean (0.1-0.4 %); two are noisy
(PN3_SID236 on plate 3, PN5_SID429 on plate 5), not a whole plate. The good donors' lines on plates 5-9 are clean.

**6. Why SNP50K does not see it.** The panel's markers are mostly not this donor's alleles, so a few percent of the donor's DNA
shows at only some of them and stays near the error rate; our markers are exactly the donor's alleles, and the veto selects
those where the lines show ALT reads. The same reads look clean on panel markers and dirty on donor markers.

## 3. Interpretation

The problem follows the family, upstream of the sequencing: both batches, separate libraries, normal BC1 pools, B73 clean,
the background specific to the donor's own alleles. The contaminating material carries this donor's alleles in much larger
amount than a BC2S3 line: an earlier generation of the same family (BC1, BC2, BC2S1) or a highly introgressed sib ear. The
user's reading: a problem in the cross, or seed from the wrong ear in the planted seed packet. A wrong ear alone would give
whole off-type plants (some lines clean genotypes of another family), not a background in every line; if each line sample is
a bulk of seedlings (the pedigree has a `BC2S3-blk` stage for these families), a few off-type seeds in the packet put a
fraction of off-type DNA into every bulk, and both batches drew from the same packets. Zx.0500_P2 shows the same at about a
third of the dose.

## Open questions

- Field and seed records of these two families: whether their BC2S3 packets or plots were near or mixed with earlier-generation
  seed of the same donor; whether batch-2 samples are bulks too.
- A per-line background measure (% non-REF reads at the donor's own sites where the line is B73) as a line QC before the
  in-silico pool and RTIGER, run over the whole population to see whether other families are affected.
- Whether the veto should act only at RTIGER's input (where it denoises) and not cut discovery
  (`agent/notes_discovery_bc1_rescue_20261007.md`).
- CRISP models every pool with 12 haplotypes (`-p 12`), the line pool (22-50 lines) included.

## Sources

Scripts and logs in `agent/union6/` (discovery-step counts, veto gain, private-site counts in `identity/`, BC1-only lane in
`bc1only_rtiger/`); SNP50K calls: zealhmm `data/zeal/rtiger_50K_calls.csv` (from zealtiger `fit_rtiger_by_donor.R`).
